Tutorial_2a

### 2024-12-01 Tutorial_2a.Rmd
correct small error
update readme and tut 2a

Tutorial 2a shows various ways to fit Bayesian binary regression models to discrete time-to-event data. The data are taken from the first experiment of Panis and Schmidt (2016), but using only the no-mask trials (with factor prime = blank, congruent, or incongruent). We perform some prior predictive checks to determine the prior distributions we will employ. We compare models using WAIC and LOO, and interpret parameter estimates for one model. 
We end by plotting the logit and cloglog link functions, and illustrate how various prior distributions on the logit and cloglog scales look on the original probability (i.e., hazard) scale.

@@

Prepare the data file, and specify priors. The skew_normal prior is set for each grand intercept on the cloglog scale, and should reflect our prior beliefs.

@@

Perform a prior predictive check to see if the binary observations generated from the specified prior distributions reflects our prior beliefs. First, sample the prior distributions using sample_prior="only".

@@

Now we can fit model M0i to get the posterior distributions.

@@

```{r save-ppc-M1c, eval=F}
ggsave("Tutorial_2_Bayesian/figures/M0c_ppc.png", width = 10, height = 8, dpi = 800)
```

>>>>

Tutorial 2a shows various ways to fit Bayesian binary regression models to discrete time-to-event data. The data are taken from the first experiment of Panis and Schmidt (2016), but using only the no-mask trials (with factor prime = blank, congruent, or incongruent). We perform some prior predictive checks to determine the prior distributions we will employ. We compare models using WAIC and LOO, and interpret parameter estimates for one model. 
We end by plotting the logit and cloglog link functions, and illustrate how various prior distributions for an intercept on the logit and cloglog scales look on the original probability (i.e., hazard) scale.

@@

Prepare the data file, and specify priors. The skew_normal prior is set for each grand intercept on the cloglog scale, and reflects our prior belief that hazard values above .5 are less likely than values below .5.

@@

### Intermezzo: Prior Predictive Check for M0i.

Perform a prior predictive check to see if the binary observations generated from the specified prior distribution(s) reflects our prior beliefs. First, sample the prior distributions using sample_prior="only".

@@

After our prior predictive check, we can fit model M0i to get the posterior distributions.

@@

```{r save-ppc-M0c, eval=F}
ggsave("Tutorial_2_Bayesian/figures/M0c_ppc.png", width = 10, height = 8, dpi = 800)
```

#
Suppose we believe apriori that hazard functions are flat.
We specify our new priors:

```{r priors-M0c-ppc2}
priors_M0c_ppc2 <- c(
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),  
  set_prior("normal(0,.001)", class = "b", coef = "period_9"), 
  set_prior("normal(0,.001)", class= "b", coef="period_9_sq"),   
  set_prior("normal(0, .001)", class = "sd"),  # set standard deviation to a small value for now
  set_prior("lkj(2)", class = "cor")  
)
```

model_M0c_prior2 <-
   brm(data = data_M0c,
       family = bernoulli(link="cloglog"),
       formula = event ~ 0 + Intercept + period_9 + period_9_sq +  
                          (0 + Intercept + period_9 + period_9_sq | pid),
       prior = priors_M0c_ppc2,
       chains = 4, cores = 4, 
       iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, 
                      step_size = 0.04, 
                      max_treedepth = 12),
       seed = 12, init = "0",
       sample_prior = "only",
       file = "Tutorial_2_Bayesian/models/model_M0c_prior2")
```

# extract predictions
for(i in 1:N_obs){
  prior_pred <- add_predicted_draws(model_M0c_prior2, 
                                    newdata = newdata_prior, 
                                    summary=F, 
                                    ndraws=200) %>% 
                mutate(obs = i)
  df_pred <- bind_rows(df_pred, prior_pred)
}

Or, suppose we believe that hazard increases with time; More specifically, we assume that cloglog-hazard increases linearly over time.

```{r priors-M0c-ppc2}
priors_M0c_ppc3 <- c(
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),  
  set_prior("normal(0.3,.001)", class = "b", coef = "period_9"), 
  set_prior("normal(0,.001)", class= "b", coef="period_9_sq"),   
  set_prior("normal(0, .001)", class = "sd"),  # set standard deviation to a small value for now
  set_prior("lkj(2)", class = "cor")  
)
```

model_M0c_prior3 <-
   brm(data = data_M0c,
       family = bernoulli(link="cloglog"),
       formula = event ~ 0 + Intercept + period_9 + period_9_sq +  
                          (0 + Intercept + period_9 + period_9_sq | pid),
       prior = priors_M0c_ppc3,
       chains = 4, cores = 4, 
       iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, 
                      step_size = 0.04, 
                      max_treedepth = 12),
       seed = 12, init = "0",
       sample_prior = "only",
       file = "Tutorial_2_Bayesian/models/model_M0c_prior3")
```

```{r ppc-model-M0c-prior, eval=F}
# Generate prior predictive hazard functions for 6 participants
# Specify
N_obs = 300 # increase for smoother results
set.seed(1)

# extract predictions
for(i in 1:N_obs){
  prior_pred <- add_predicted_draws(model_M0c_prior3, 
                                    newdata = newdata_prior, 
                                    summary=F, 
                                    ndraws=20) %>% 
                mutate(obs = i)
  df_pred <- bind_rows(df_pred, prior_pred)
}

# plot 200 draws per participant, highlighting three
sel = sample(1:20,3,replace=F)

ggplot(df_pred_haz %>% 
   filter(.draw <= 20), aes(x=period_9, y=pred_haz, group=.draw)) +
   geom_line(color="black") +
   # highlight
   geom_line(data=df_pred_haz %>% filter(.draw ==sel[1]), 
             color="red", linewidth=2) +
   geom_line(data=df_pred_haz %>% filter(.draw == sel[2]), 
            color="green", linewidth=2) +
   geom_line(data=df_pred_haz %>% filter(.draw == sel[3]), 
          color="blue", linewidth=2) +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(limits = c(-3,6), breaks = c(-3:6)) +
  facet_wrap(~pid)
```

Or, suppose we believe that hazard first increases with time, and then decreases;

```{r priors-M0c-ppc2}
priors_M0c_ppc3 <- c(
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),  
  set_prior("normal(0.6,.2)", class = "b", coef = "period_9"), 
  set_prior("normal(-0,.09)", class= "b", coef="period_9_sq"),   
  set_prior("normal(0, .1)", class = "sd"),  # set standard deviation to a small value for now
  set_prior("lkj(2)", class = "cor")  
)
```

model_M0c_prior3 <-
   brm(data = data_M0c,
       family = bernoulli(link="cloglog"),
       formula = event ~ 0 + Intercept + period_9 + period_9_sq +  
                          (0 + Intercept + period_9 + period_9_sq | pid),
       prior = priors_M0c_ppc3,
       chains = 4, cores = 4, 
       iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, 
                      step_size = 0.04, 
                      max_treedepth = 12),
       seed = 12, init = "0",
       sample_prior = "only",
       file = "Tutorial_2_Bayesian/models/model_M0c_prior3")
```

```{r ppc-model-M0c-prior, eval=F}
# Generate prior predictive hazard functions for 6 participants
# Specify
N_obs = 100 # increase for smoother results
N_samples = 8000
set.seed(1)

# extract predictions
for(i in 1:N_obs){
  prior_pred <- add_predicted_draws(model_M0c_prior3, 
                                    newdata = newdata_prior, 
                                    summary=F, 
                                    ndraws=N_samples) %>% 
                mutate(obs = i)
  df_pred <- bind_rows(df_pred, prior_pred)
}

# plot 200 draws per participant, highlighting three
sel = sample(1:N_samples,3,replace=F)

ggplot(df_pred_haz %>% 
   filter(.draw <= N_samples), aes(x=period_9, y=pred_haz, group=.draw)) +
   geom_line(color="black") +
   # highlight
   geom_line(data=df_pred_haz %>% filter(.draw ==sel[1]), 
             color="red", linewidth=2) +
   geom_line(data=df_pred_haz %>% filter(.draw == sel[2]), 
            color="green", linewidth=2) +
   geom_line(data=df_pred_haz %>% filter(.draw == sel[3]), 
          color="blue", linewidth=2) +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(limits = c(-3,6), breaks = c(-3:6)) +
  facet_wrap(~pid)
```

Or, check settings.

```{r priors-M0c-ppc2}
priors_M0c_ppc3 <- c(
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),  
  set_prior("normal(0.1,.1)", class = "b", coef = "period_9"), 
  set_prior("normal(0,.06)", class= "b", coef="period_9_sq"),   
  set_prior("normal(0, .1)", class = "sd"),  # set standard deviation to a small value for now
  set_prior("lkj(2)", class = "cor")  
)
```

model_M0c_prior3 <-
   brm(data = data_M0c,
       family = bernoulli(link="cloglog"),
       formula = event ~ 0 + Intercept + period_9 + period_9_sq +  
                          (0 + Intercept + period_9 + period_9_sq | pid),
       prior = priors_M0c_ppc3,
       chains = 4, cores = 4, 
       iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, 
                      step_size = 0.04, 
                      max_treedepth = 12),
       seed = 12, init = "0",
       sample_prior = "only",
       file = "Tutorial_2_Bayesian/models/model_M0c_prior3")
```

```{r ppc-model-M0c-prior, eval=F}
# Generate prior predictive hazard functions for 6 participants
# Specify
N_obs = 100 # increase for smoother results
N_samples = 8000
set.seed(1)

# extract predictions
for(i in 1:N_obs){
  prior_pred <- add_predicted_draws(model_M0c_prior3, 
                                    newdata = newdata_prior, 
                                    summary=F, 
                                    ndraws=N_samples) %>% 
                mutate(obs = i)
  df_pred <- bind_rows(df_pred, prior_pred)
}

# plot 200 draws per participant, highlighting three
sel = sample(1:N_samples,3,replace=F)

ggplot(df_pred_haz %>% 
   filter(.draw <= N_samples), aes(x=period_9, y=pred_haz, group=.draw)) +
   geom_line(color="black") +
   # highlight
   geom_line(data=df_pred_haz %>% filter(.draw ==sel[1]), 
             color="red", linewidth=2) +
   geom_line(data=df_pred_haz %>% filter(.draw == sel[2]), 
            color="green", linewidth=2) +
   geom_line(data=df_pred_haz %>% filter(.draw == sel[3]), 
          color="blue", linewidth=2) +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(limits = c(-3,6), breaks = c(-3:6)) +
  facet_wrap(~pid)
```

#
